% Neural Surrogate Modeling for Snake Robot Locomotion
% Research Report — Graduation Project Deliverable

\documentclass[11pt]{article}

% ---------------------------------------------------------------------------
% Font and encoding
% ---------------------------------------------------------------------------
\usepackage{palatino}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}

% ---------------------------------------------------------------------------
% Page layout
% ---------------------------------------------------------------------------
\usepackage[top=1in, bottom=1in, left=1.25in, right=1.25in]{geometry}
\usepackage{setspace}
\setstretch{1.1}
\setlength{\parindent}{1em}
\setlength{\parskip}{0.4em}

% ---------------------------------------------------------------------------
% Math
% ---------------------------------------------------------------------------
\usepackage{amsmath}
\usepackage{amssymb}

% ---------------------------------------------------------------------------
% Figures and tables
% ---------------------------------------------------------------------------
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{float}
\usepackage{booktabs}
\graphicspath{{../figures/}}

% ---------------------------------------------------------------------------
% Colors
% ---------------------------------------------------------------------------
\usepackage{xcolor}

% ---------------------------------------------------------------------------
% Lists
% ---------------------------------------------------------------------------
\usepackage{enumitem}
\setlist[itemize]{nosep, topsep=0.3em, itemsep=0.2em}

% ---------------------------------------------------------------------------
% Algorithms
% ---------------------------------------------------------------------------
\usepackage{algorithm}
\usepackage{algorithmic}

% ---------------------------------------------------------------------------
% Section spacing
% ---------------------------------------------------------------------------
\usepackage{titlesec}
\titlespacing*{\section}{0pt}{1.2em}{0.6em}
\titlespacing*{\subsection}{0pt}{0.9em}{0.3em}
\titlespacing*{\subsubsection}{0pt}{0.7em}{0.2em}

% ---------------------------------------------------------------------------
% Header and footer
% ---------------------------------------------------------------------------
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhf{}
\lhead{\small Neural Surrogate for Snake Robot RL}
\rfoot{\thepage}
\renewcommand{\headrulewidth}{0.4pt}

% ---------------------------------------------------------------------------
% Table of contents and section depth
% ---------------------------------------------------------------------------
\setcounter{tocdepth}{3}
\setcounter{secnumdepth}{3}

% ---------------------------------------------------------------------------
% Placeholder macros
% ---------------------------------------------------------------------------
\newcommand{\placeholder}[1]{\textcolor{gray}{\textit{[Placeholder: #1]}}}
\newcommand{\mytodo}[1]{\textcolor{red}{\textbf{[TODO: #1]}}}

% ---------------------------------------------------------------------------
% Bibliography (natbib — before hyperref)
% ---------------------------------------------------------------------------
\usepackage[round, sort, authoryear]{natbib}

% ---------------------------------------------------------------------------
% Hyperref and microtype — before cleveref
% ---------------------------------------------------------------------------
\usepackage[hidelinks]{hyperref}
\usepackage{microtype}

% ---------------------------------------------------------------------------
% Cross-references (cleveref — MUST be loaded after hyperref)
% ---------------------------------------------------------------------------
\usepackage{cleveref}

% ---------------------------------------------------------------------------
% Title
% ---------------------------------------------------------------------------
\title{%
  \textbf{Neural Surrogate Modeling for Snake Robot Locomotion:}\\[0.3em]
  \large Faster Reinforcement Learning via Learned Cosserat Rod Dynamics%
}
\author{%
  Qifan Wen\\
  \small Electrical and Computer Engineering\\
  \small The Ohio State University\\
  \small \texttt{wen.679@buckeyemail.osu.edu}\\[0.6em]
  \small \textit{Advisor:} Prof.\ Abhishek Gupta\quad\small\texttt{gupta.706@osu.edu}%
}
\date{}

% ===========================================================================
\begin{document}
% ===========================================================================

\maketitle

% ---------------------------------------------------------------------------
\begin{abstract}
\placeholder{Write abstract after all experimental results are available.}
\end{abstract}

\tableofcontents
\newpage

% ---------------------------------------------------------------------------
% Notation table
% ---------------------------------------------------------------------------
\section*{Notation}
\label{sec:notation}
\addcontentsline{toc}{section}{Notation}

\begin{table}[H]
  \centering
  \small
  \begin{tabular}{lll}
    \toprule
    Symbol & Dimension & Meaning \\
    \midrule
    $\mathbf{s}_t$ & $\mathbb{R}^{124}$ & Rod state vector at time step $t$ \\
    $\mathbf{a}_t$ & $[-1,1]^5$ & Normalized CPG action vector \\
    $T$ & $\mathbb{R}^{124} \times \mathbb{R}^5 \to \mathbb{R}^{124}$ & Ground-truth transition operator (PyElastica) \\
    $f_\theta$ & $\mathbb{R}^{189} \to \mathbb{R}^{124}$ & Neural surrogate model with parameters $\theta$ \\
    $\mathbf{z}_t$ & $\mathbb{R}^{189}$ & Surrogate input (state $\|$ action $\|$ phase encoding) \\
    $\boldsymbol{\phi}_e$ & $\mathbb{R}^3$ & Per-element CPG phase triplet for element $e$ \\
    $\mathbf{x}(s,t)$ & $\mathbb{R}^3$ & Rod centerline position at arc length $s$ \\
    $\mathbf{d}(s,t)$ & $\mathbb{R}^3$ & Material director frame (orientation) \\
    $\mathbf{F}$ & $\mathbb{R}^3$ & Internal force (shear and tension) \\
    $\mathbf{M}$ & $\mathbb{R}^3$ & Internal moment (bending and torsion) \\
    $\kappa_e^{\text{target}}$ & $\mathbb{R}$ & Target curvature at element $e$ \\
    $\mu_\Delta, \sigma_\Delta$ & $\mathbb{R}^{124}$ & Training-set mean and std of state deltas \\
    $N$ & 21 & Number of discretization nodes \\
    $N_e$ & 20 & Number of discretization elements \\
    \bottomrule
  \end{tabular}
\end{table}

% ---------------------------------------------------------------------------
\section{Introduction}
\label{sec:intro}
% ---------------------------------------------------------------------------

Soft and continuum robots present a fundamental challenge for reinforcement learning: their dynamics are governed by distributed elasticity rather than rigid joints, and the resulting equations of motion are expensive to simulate \citep{Bing2019}.
A snake robot modeled as a Cosserat rod requires the resolution of elastic body dynamics, anisotropic ground friction, and internally generated muscle moments at every simulation substep.
PyElastica \citep{Naughton2021}, an open-source Cosserat rod simulator, provides this capability through a symplectic Euler integrator with hundreds of internal substeps per control step.
The simulation is physically faithful, but computationally demanding.

On a 48-core server with 16 parallel workers, each environment step requires approximately 46\,ms, yielding a throughput of roughly 350 steps per second.
A proximal policy optimization (PPO) training run requiring $2 \times 10^6$ environment interactions therefore takes approximately 36 hours of wall-clock time.
Practical RL experimentation --- where multiple runs with different hyperparameters or random seeds are needed --- multiplies this cost proportionally.
This computational bottleneck is the central motivation for the present work.

The key observation underlying the surrogate approach is that the mapping from the current state and action to the next state --- the transition operator $T : (\mathbf{s}_t, \mathbf{a}_t) \mapsto \mathbf{s}_{t+1}$ --- is a deterministic function that can be approximated by a neural network.
If a learned surrogate $f_\theta \approx T$ can be evaluated in a single GPU forward pass rather than through hundreds of symplectic Euler substeps, the per-step cost drops from ${\sim}46$\,ms to ${\sim}0.001$\,ms.
This represents a projected speedup of four to five orders of magnitude.
Such a speedup, if achieved at acceptable prediction accuracy, would reduce RL training from a multi-day commitment to a task completable in minutes.

A three-stage pipeline is presented in this report.
First, a parallel data collection system generates a large, diverse dataset of transition pairs $(\mathbf{s}_t, \mathbf{a}_t, \mathbf{s}_{t+1})$ from 16 independent PyElastica workers using Sobol quasi-random action sampling and state perturbation (\cref{sec:methods:datacollection}).
Second, a feedforward MLP surrogate is trained to predict the state delta $\mathbf{s}_{t+1} - \mathbf{s}_t$ from a 189-dimensional input comprising the 124-dimensional rod state, the 5-dimensional CPG action, and a 60-dimensional per-element phase encoding that provides the model with the exact phase of the serpenoid wave at each rod element (\cref{sec:methods:encoding,sec:methods:architecture}).
Third, the trained surrogate serves as a drop-in replacement for the PyElastica environment during PPO training, with a direct Elastica RL baseline providing the controlled comparison that isolates the policy gap attributable to surrogate modeling error (\cref{sec:experiments}).

The remainder of this report is organized as follows.
\Cref{sec:background} introduces the Cosserat rod dynamics, the resistive force theory contact model, the central pattern generator actuation scheme, and the PyElastica simulator.
\Cref{sec:related} reviews prior work on neural surrogates for soft robot dynamics and reinforcement learning with learned simulators.
\Cref{sec:methods} describes the data collection pipeline, state and action representations, per-element phase encoding, surrogate architecture, and the two-phase training procedure.
\Cref{sec:experiments} presents surrogate prediction accuracy, RL training performance, and the comparison with the direct Elastica baseline.
\Cref{sec:discussion} reflects on the physics calibration challenges, architecture experiments, and engineering lessons learned from the data collection pipeline.
\Cref{sec:conclusion} summarizes the findings and outlines directions for future work.

% ---------------------------------------------------------------------------
\section{Background}
\label{sec:background}
% ---------------------------------------------------------------------------

\subsection{Cosserat Rod Dynamics}
\label{sec:background:cosserat}

The snake robot body is modeled as a Cosserat rod --- a one-dimensional elastic continuum that supports shear, extension, bending, and torsion along its arc length \citep{Till2019}.
The Cosserat formulation captures the distributed compliance of the rod body, making it appropriate for soft and continuum robots whose deformation is governed by material elasticity rather than joint constraints.
The governing equations, expressed per unit arc length $s$, enforce the conservation of linear and angular momentum:

\begin{equation}
  \rho A \frac{\partial^2 \mathbf{x}}{\partial t^2}
    = \frac{\partial \mathbf{F}}{\partial s} + \mathbf{f}_{\text{ext}}
  \label{eq:linear-momentum}
\end{equation}

\begin{equation}
  \rho \mathbf{I} \frac{\partial^2 \mathbf{d}}{\partial t^2}
    = \frac{\partial \mathbf{M}}{\partial s}
      + \mathbf{m}_{\text{ext}}
      + \frac{\partial \mathbf{x}}{\partial s} \times \mathbf{F}
  \label{eq:angular-momentum}
\end{equation}

\smallskip
\noindent where
\begin{tabular}{@{}>{$}l<{$} @{\quad} l@{}}
  \mathbf{x}(s,t) & rod centerline position (m) \\
  \mathbf{d}(s,t) & material director frame (local orientation) \\
  \mathbf{F}       & internal force (shear + tension, via constitutive law) \\
  \mathbf{M}       & internal moment (bending + torsion) \\
  \mathbf{f}_{\text{ext}} & external force per unit length \\
  \mathbf{m}_{\text{ext}} & external moment per unit length \\
  \rho A           & linear mass density \\
  \rho \mathbf{I}  & rotational inertia per unit length \\
\end{tabular}

The external force $\mathbf{f}_{\text{ext}}$ captures the ground contact forces described in \cref{sec:background:rft}, while $\mathbf{m}_{\text{ext}}$ represents the muscle moments generated by the CPG actuator (\cref{sec:background:cpg}).

Although the full Cosserat model operates in three dimensions, snake locomotion in this work is constrained to the horizontal plane.
All $z$-components of position and velocity are verified to be identically zero, so the system reduces to a two-dimensional projection without loss of physical content.

The PDEs in \cref{eq:linear-momentum,eq:angular-momentum} are discretized spatially into $N = 21$ nodes and $N_e = 20$ elements, producing a first-order ODE system that is integrated forward in time by a symplectic Euler stepper.
The resulting discrete state vector $\mathbf{s}_t \in \mathbb{R}^{124}$ is defined in \cref{tab:state-vector}.

\begin{table}[H]
  \centering
  \small
  \caption{Layout of the 124-dimensional rod state vector $\mathbf{s}_t$.}
  \label{tab:state-vector}
  \begin{tabular}{llrl}
    \toprule
    Index slice & Variable & Count & Physical meaning \\
    \midrule
    \texttt{[0:21]}    & $x_i$          & 21 & Node $x$-positions (m)                         \\
    \texttt{[21:42]}   & $y_i$          & 21 & Node $y$-positions (m)                         \\
    \texttt{[42:63]}   & $\dot{x}_i$    & 21 & Node $x$-velocities (m\,s$^{-1}$)             \\
    \texttt{[63:84]}   & $\dot{y}_i$    & 21 & Node $y$-velocities (m\,s$^{-1}$)             \\
    \texttt{[84:104]}  & $\psi_e$       & 20 & Element yaw angle (rad)                        \\
    \texttt{[104:124]} & $\omega_{z,e}$ & 20 & Element angular velocity about $z$-axis (rad\,s$^{-1}$) \\
    \bottomrule
  \end{tabular}
\end{table}

This 124-dimensional vector is the Markov state used for both surrogate training and RL policy input.
The lower-dimensional observation space (14 floats derived from FFT compression of node positions) that the RL policy receives is not Markov: identical observations can correspond to different underlying rod configurations and velocity profiles.
As a consequence, the surrogate must operate on the full 124-dimensional state to produce correct multi-step rollouts.

\subsection{Resistive Force Theory Contact Model}
\label{sec:background:rft}

Ground contact forces are modeled using Resistive Force Theory (RFT), a slender-body approximation in which the force at each rod element depends only on the local velocity resolved into tangential and normal components relative to the element centerline \citep{Till2019}.
This formulation avoids the computational expense of a full contact optimization at each substep while capturing the anisotropic friction responsible for serpentine propulsion.

The drag forces per unit length on element $e$ are given by:

\begin{equation}
  f_t = -c_t \, v_t
  \label{eq:rft-tangential}
\end{equation}

\begin{equation}
  f_n = -c_n \, v_n
  \label{eq:rft-normal}
\end{equation}

\smallskip
\noindent where
\begin{tabular}{@{}>{$}l<{$} @{\quad} l@{}}
  v_t & velocity component tangent to the element \\
  v_n & velocity component normal to the element \\
  c_t & tangential drag coefficient \\
  c_n & normal drag coefficient \\
\end{tabular}

The values used in this work are $c_t = 0.01$ and $c_n = 0.05$, giving a normal-to-tangential ratio of five.
The anisotropy $c_n > c_t$ is the physical mechanism enabling net forward motion: lateral slip is resisted more strongly than axial slip, so the reaction forces from a traveling curvature wave produce a net thrust in the direction of wave propagation.

\subsection{Central Pattern Generator Actuation}
\label{sec:background:cpg}

Locomotion is produced by a traveling curvature wave imposed along the rod body.
A Central Pattern Generator (CPG) controller prescribes a target curvature at each element; an elastic restoring force within the simulator then drives the rod toward the target, producing the serpentine gait.
The target curvature at element $i$ is:

\begin{equation}
  \kappa_i^{\text{target}}
    = A \sin\!\bigl(2\pi k \cdot s_i + 2\pi f \cdot t + \phi_0\bigr) + b
  \label{eq:cpg-curvature}
\end{equation}

\smallskip
\noindent where
\begin{tabular}{@{}>{$}l<{$} @{\quad} l@{}}
  s_i    & normalized arc-length position of element $i$ ($s_i \in [0,1]$) \\
  A      & wave amplitude (rad\,m$^{-1}$) \\
  k      & wave number (wavelengths per body length) \\
  f      & oscillation frequency (Hz) \\
  \phi_0 & initial phase offset (rad) \\
  b      & constant turn bias (rad\,m$^{-1}$) \\
\end{tabular}

Together, these form the five-dimensional action vector:

\begin{equation}
  \mathbf{a} = (A,\; f,\; k,\; \phi_0,\; b) \in \mathbb{R}^5
  \label{eq:action-vector}
\end{equation}

The physical ranges are $A \in [0, 5]$, $f \in [0.5, 3.0]$\,Hz, $k \in [0.5, 3.5]$, $\phi_0 \in [0, 2\pi]$, and $b \in [-2, 2]$.
The RL policy outputs a normalized action $\mathbf{a} \in [-1,1]^5$ that is linearly mapped to these physical ranges at each control step.

\subsection{PyElastica Simulator}
\label{sec:background:elastica}

PyElastica \citep{Naughton2021} is an open-source Python framework for compliant mechanics simulation based on Cosserat rod theory.
The discretized equations of motion from \cref{sec:background:cosserat} are integrated using a symplectic Euler scheme, which provides numerical stability for stiff elastic dynamics without the cost of implicit methods.
Each RL control step of duration $\Delta t_{\text{ctrl}}$ executes $n_{\text{sub}}$ internal physics substeps, defining the ground-truth transition operator:

\begin{equation}
  T : (\mathbf{s}_t, \mathbf{a}_t) \mapsto \mathbf{s}_{t+1}
  \label{eq:transition-operator}
\end{equation}

A single evaluation of $T$ requires approximately 46\,ms when 16 parallel workers are active on a 48-core CPU server.
The resulting throughput of ${\sim}350$ steps per second means that $2 \times 10^6$ environment interactions --- a typical PPO training budget --- require approximately 36 hours.
The surrogate model $f_\theta \approx T$ replaces these hundreds of substeps with a single GPU forward pass, projecting a throughput improvement of four to five orders of magnitude.

% ---------------------------------------------------------------------------
\section{Related Work}
\label{sec:related}
% ---------------------------------------------------------------------------

\subsection{Neural Surrogates for Soft Robot Dynamics}
\label{sec:related:surrogates}

The use of neural networks to approximate the dynamics of elastic and soft robots has been investigated as a means of bypassing the computational cost of first-principles simulation.
Several results are directly relevant to the Cosserat rod setting considered here.

\citet{Stolzle2025} propose a domain-decoupled physics-informed neural network (DD-PINN) for Cosserat rod dynamics.
By decoupling the time domain from the feedforward network, gradients are computed in closed form rather than through automatic differentiation over the ODE integrator.
The resulting model achieves a $44{,}000\times$ speedup over numerical Cosserat rod integration, with end-effector position errors below 3\,mm (2.3\% of actuator length) across the full 72-dimensional rod state.
This performance enables nonlinear model-predictive control at 70\,Hz on a GPU --- a capability that is infeasible with direct numerical integration.

A hybrid approach is taken in KNODE-Cosserat \citep{Hsieh2024}, where a simplified first-principles Cosserat rod model serves as the backbone and a neural ODE residual term corrects the systematic discrepancy between the simplified physics and the true behavior.
Training data for this model is collected from PyElastica simulations --- a pipeline analogous to the one used here.
The hybrid model achieves a 58.7\% accuracy improvement over the physics-only model on long-horizon trajectories, demonstrating that even a coarse physics prior reduces the learning burden on the neural correction term.

The most closely related prior work is SoRoLEX \citep{SoRoLEX2024}, which trains an LSTM forward dynamics model on soft robot simulator data and deploys it as a surrogate environment for PPO training with JAX-based GPU parallelization.
The pipeline --- collect trajectories from a high-fidelity simulator, train a neural dynamics model, run large-scale RL in the learned environment --- is identical in structure to the approach presented here.
The differences lie in the architecture (MLP rather than LSTM), the prediction target (the full 124-dimensional Cosserat rod state rather than a task-space observation), and the explicit use of a multi-step rollout loss to limit autoregressive error accumulation.

Physics-informed approaches have also been applied to articulated soft robots.
\citet{PINNSoftRobot2025} demonstrate a $467\times$ speedup over first-principles models with MPC at 47\,Hz, requiring as little as a single training dataset.
This illustrates the data efficiency advantages that follow from incorporating physical structure into the surrogate architecture.

These results collectively establish that neural surrogates for Cosserat rod dynamics are feasible and capable of sustaining the multi-step rollout accuracy required for closed-loop control.
The present work follows the data-driven approach of SoRoLEX, extended with per-element CPG phase encoding and a two-phase training schedule that combines single-step MSE with multi-step rollout loss.

\subsection{Reinforcement Learning with Learned Simulators}
\label{sec:related:rl}

The idea of replacing expensive environment interactions with a learned dynamics model originates with the Dyna architecture \citep{Sutton1991} and has since been developed into a practical methodology by several influential results.

\citet{Janner2019} demonstrate in Model-Based Policy Optimization (MBPO) that an ensemble of learned dynamics models can generate short synthetic rollouts from real environment states.
Policy optimization on these rollouts matches model-free asymptotic performance while requiring approximately five times fewer real-environment interactions.
The central insight --- that short-horizon model predictions limit error accumulation while still providing useful policy gradient signal --- directly motivates the surrogate-based RL approach taken here.

The world model paradigm extends this idea by learning compact latent dynamics alongside the policy.
DreamerV3 \citep{Hafner2023} trains a recurrent state-space model entirely from sensory observations and sustains imagination-based RL across more than 150 diverse tasks without task-specific tuning.
This result illustrates that surrogate quality is the primary bottleneck for downstream policy performance.
In the present work, the accuracy of the Cosserat rod surrogate --- quantified in \cref{sec:experiments:accuracy} --- directly determines the quality of policies trained in the surrogate environment.
The Phase~8 comparison with a direct PyElastica RL baseline isolates the policy gap attributable to surrogate error.

Graph-structured alternatives to MLP surrogates have also been explored for deformable body simulation.
\citet{SanchezGonzalez2020} introduce a particle-based graph network simulator that achieves up to $1000\times$ speedups for fluid, rigid, and deformable material dynamics.
\citet{Pfaff2021} extend this approach to unstructured simulation meshes with MeshGraphNets, which generalize across different mesh resolutions.
For the snake robot considered here, however, the rod is discretized into only 20 elements --- a graph small enough that a dense MLP achieves comparable expressive power at substantially lower implementation cost.
Graph-structured surrogates would become advantageous for rods with significantly more nodes or for multi-rod assemblies interacting through contact.

% ---------------------------------------------------------------------------
\section{Methods}
\label{sec:methods}
% ---------------------------------------------------------------------------

\subsection{Data Collection Pipeline}
\label{sec:methods:datacollection}

Training data for the surrogate model is generated by running 16 independent worker processes, each maintaining its own PyElastica Cosserat rod simulation instance.
Actions are sampled from the 5-dimensional CPG action space using a Sobol quasi-random sequence, which distributes points more uniformly across all dimensions than pseudo-random sampling.
To populate diverse regions of the state space, 30\% of episodes begin with a state perturbation: random offsets are applied to node positions, and angular velocity noise is injected into the rod elements with $\sigma_\omega = 1.5$\,rad/s (see \cref{sec:discussion:pipeline} for the rationale behind this value).
Within each episode, 50\% of actions are drawn uniformly at random and the remaining 50\% follow the Sobol sequence.

\begin{algorithm}[H]
\caption{Parallel data collection for surrogate training}
\label{alg:data-collection}
\begin{algorithmic}[1]
  \REQUIRE Number of workers $W = 16$, Sobol sequence $\mathcal{S}$, perturbation fraction $p = 0.3$
  \ENSURE Dataset $\mathcal{D} = \{(\mathbf{s}_t, \mathbf{a}_t, \mathbf{s}_{t+1})\}$
  \FOR{each worker $w = 1, \ldots, W$ in parallel}
    \LOOP
      \STATE Initialize rod; with probability $p$, perturb state
      \STATE Sample action $\mathbf{a}$ from $\mathcal{S}$ (50\%) or uniform (50\%)
      \STATE Execute one RL step: $\mathbf{s}_{t+1} \leftarrow T(\mathbf{s}_t, \mathbf{a}_t)$
      \STATE Store $(\mathbf{s}_t, \mathbf{a}_t, \mathbf{s}_{t+1}, t_{\text{start}})$ to batch file
    \ENDLOOP
  \ENDFOR
\end{algorithmic}
\end{algorithm}

The collection format stores one transition per row: the current state $\mathbf{s}_t$, the action $\mathbf{a}_t$, the next state $\mathbf{s}_{t+1}$, the episode start time $t_{\text{start}}$, and optionally the internal and external forces and torques acting on the rod.
At approximately 2055 bytes per transition including forces, the target dataset size of 10\,GB yields roughly 5 million transition pairs.
A health monitor tracks the output rate of each worker and respawns any worker that stalls after initialization, ensuring sustained throughput across multi-hour collection runs.

\subsection{State and Action Representation}
\label{sec:methods:representation}

The full 124-dimensional rod state vector $\mathbf{s}_t \in \mathbb{R}^{124}$, defined in \cref{tab:state-vector}, serves as the Markov state for both surrogate training and RL.
The 2D projection from the 3D Cosserat state is exact --- $z$-components are identically zero --- so no physical content is lost.

The action vector $\mathbf{a}_t \in [-1,1]^5$ is linearly mapped to the physical CPG parameter ranges defined in \cref{sec:background:cpg}:

\begin{equation}
  \mathbf{a}_{\text{phys}} = \frac{1}{2}\bigl[(\mathbf{a}_t + 1) \odot (\mathbf{a}_{\max} - \mathbf{a}_{\min}) + 2\,\mathbf{a}_{\min}\bigr]
  \label{eq:action-mapping}
\end{equation}

\smallskip
\noindent where
\begin{tabular}{@{}>{$}l<{$} @{\quad} l@{}}
  \mathbf{a}_t           & normalized action from the RL policy ($\in [-1,1]^5$) \\
  \mathbf{a}_{\text{phys}} & action in physical CPG parameter units \\
  \mathbf{a}_{\max},\, \mathbf{a}_{\min} & upper and lower bounds of each CPG parameter \\
  \odot                  & element-wise (Hadamard) product \\
\end{tabular}

Denormalization to physical units occurs only inside the PyElastica environment; the surrogate receives the normalized action $\mathbf{a}_t$ directly.

\subsection{Per-Element CPG Phase Encoding}
\label{sec:methods:encoding}

The target curvature at element $e$ depends on both the global oscillation phase $2\pi f t$ and the element's arc-length position $s_e$ through the serpenoid wave (\cref{eq:cpg-curvature}).
Elements at opposite ends of the rod can be at opposite phases of the curvature wave at the same instant.
An early version of the surrogate (V1) used a global phase encoding $[\sin(2\pi f t), \cos(2\pi f t)] \in \mathbb{R}^2$ appended once to the input, requiring the model to reconstruct per-element phase offsets implicitly through shared MLP weights.
This imposed a significant learning burden and contributed to poor angular velocity predictions (R$^2 = 0.23$).

The V2 encoding provides the per-element phase explicitly.
For element $e \in \{1, \ldots, N_e\}$, an effective time is computed as:

\begin{equation}
  t_e = \frac{k \cdot s_e}{f}
  \label{eq:effective-time}
\end{equation}

From this, a phase triplet is constructed:

\begin{equation}
  \boldsymbol{\phi}_e
    = \bigl(\sin(2\pi f \cdot t_e),\;
            \cos(2\pi f \cdot t_e),\;
            \kappa_e^{\text{actual}}\bigr)
  \label{eq:per-element-phase}
\end{equation}

\smallskip
\noindent where
\begin{tabular}{@{}>{$}l<{$} @{\quad} l@{}}
  t_e & effective time incorporating spatial phase offset at element $e$ \\
  k   & wave number (wavelengths per body length) \\
  s_e & normalized arc-length position of element $e$ \\
  f   & oscillation frequency (Hz) \\
  \sin(2\pi f \cdot t_e),\; \cos(2\pi f \cdot t_e) & exact phase of the serpenoid wave at element $e$ \\
  \kappa_e^{\text{actual}} & instantaneous curvature at element $e$ \\
\end{tabular}

Stacking $\boldsymbol{\phi}_e$ for all $N_e = 20$ elements yields a 60-dimensional phase feature vector.
These features are computed on-the-fly during dataset loading from the stored action and episode start time $t_{\text{start}}$, so batch files contain only rod states, actions, and $t_{\text{start}}$ without pre-computed phase data.

The full surrogate input concatenates the z-score normalized rod state $\bar{\mathbf{s}}_t$, the normalized action $\mathbf{a}_t$, and the per-element phase features:

\begin{equation}
  \mathbf{z}_t
    = \bigl[\bar{\mathbf{s}}_t \;\|\; \mathbf{a}_t \;\|\;
            \boldsymbol{\phi}_1, \ldots, \boldsymbol{\phi}_{N_e}\bigr]
    \in \mathbb{R}^{189}
  \label{eq:surrogate-input}
\end{equation}

\smallskip
\noindent where
\begin{tabular}{@{}>{$}l<{$} @{\quad} l@{}}
  \bar{\mathbf{s}}_t & z-score normalized state: $(\mathbf{s}_t - \boldsymbol{\mu}_s) / \boldsymbol{\sigma}_s$ \\
  \mathbf{a}_t & normalized CPG action ($\in [-1,1]^5$) \\
  \boldsymbol{\phi}_1, \ldots, \boldsymbol{\phi}_{N_e} & per-element phase triplets ($60$ dimensions total) \\
  \boldsymbol{\mu}_s,\, \boldsymbol{\sigma}_s & training-set per-feature mean and standard deviation \\
\end{tabular}

The 189 dimensions decompose as $124 + 5 + 60$.

\subsection{Surrogate Model Architecture}
\label{sec:methods:architecture}

The surrogate is a feedforward MLP that maps the 189-dimensional input to a 124-dimensional normalized state delta.
The predicted next state is recovered through an affine reconstruction:

\begin{equation}
  f_\theta : \mathbf{z}_t \mapsto \overline{\Delta\mathbf{s}}
  \label{eq:surrogate-forward}
\end{equation}

\begin{equation}
  \hat{\mathbf{s}}_{t+1}
    = \mathbf{s}_t
      + \boldsymbol{\sigma}_\Delta \odot f_\theta(\mathbf{z}_t)
      + \boldsymbol{\mu}_\Delta
  \label{eq:next-state-reconstruction}
\end{equation}

\smallskip
\noindent where
\begin{tabular}{@{}>{$}l<{$} @{\quad} l@{}}
  f_\theta & neural surrogate with parameters $\theta$ \\
  \overline{\Delta\mathbf{s}} & predicted normalized state delta \\
  \boldsymbol{\mu}_\Delta & training-set mean of $(\mathbf{s}_{t+1} - \mathbf{s}_t)$ \\
  \boldsymbol{\sigma}_\Delta & training-set std of $(\mathbf{s}_{t+1} - \mathbf{s}_t)$ \\
  \odot & element-wise (Hadamard) product \\
\end{tabular}

The statistics $\boldsymbol{\mu}_\Delta, \boldsymbol{\sigma}_\Delta \in \mathbb{R}^{124}$ are computed once before training begins.

Delta prediction offers two advantages over predicting the absolute next state.
First, per-step state changes are small relative to absolute values --- position deltas are typically on the order of centimeters per control step while absolute positions span meters --- so the target distribution is more concentrated and easier to fit.
Second, when $f_\theta(\mathbf{z}_t) \approx \mathbf{0}$, the reconstruction yields $\hat{\mathbf{s}}_{t+1} \approx \mathbf{s}_t + \boldsymbol{\mu}_\Delta$, providing a reasonable default (the mean empirical transition) that acts as an implicit residual connection.

The network architecture is:

\begin{equation}
  \mathbb{R}^{189}
    \xrightarrow{\text{Linear}}
    \bigl[\text{Linear}(512) \to \text{LayerNorm} \to \text{SiLU}\bigr]^{\times 3}
    \xrightarrow{\text{Linear}}
    \mathbb{R}^{124}
  \label{eq:architecture}
\end{equation}

The output layer is initialized to zero so that the model starts by predicting zero delta.
LayerNorm is applied after each linear transformation (before the SiLU activation) to stabilize training dynamics.
The total number of trainable parameters is approximately 659K.

\subsection{Training Procedure}
\label{sec:methods:training}

Training is organized in two phases that introduce the rollout loss gradually to avoid instability during early optimization.

\subsubsection{Phase 1: Single-Step MSE (Epochs 1--20)}
\label{sec:methods:training:phase1}

The model is trained to minimize the mean squared error on the normalized state delta:

\begin{equation}
  \mathcal{L}_{\text{single}}
    = \frac{1}{N} \sum_{i=1}^{N}
      \bigl\| f_\theta(\mathbf{z}_t^{(i)})
             - \overline{\Delta\mathbf{s}}_{\text{true}}^{(i)} \bigr\|^2
  \label{eq:single-loss}
\end{equation}

\smallskip
\noindent where
\begin{tabular}{@{}>{$}l<{$} @{\quad} l@{}}
  N & batch size \\
  f_\theta(\mathbf{z}_t^{(i)}) & predicted normalized state delta for sample $i$ \\
  \overline{\Delta\mathbf{s}}_{\text{true}}^{(i)} & ground-truth normalized state delta for sample $i$ \\
\end{tabular}

This phase establishes accurate one-step predictions before the more expensive rollout computation is introduced.

\subsubsection{Phase 2: Combined Loss with Rollout (Epoch 20 Onward)}
\label{sec:methods:training:phase2}

An 8-step autoregressive rollout loss is added to penalize error accumulation:

\begin{equation}
  \mathcal{L}
    = \mathcal{L}_{\text{single}}
      + \lambda_r \cdot \mathcal{L}_{\text{rollout}}
  \label{eq:combined-loss}
\end{equation}

\begin{equation}
  \mathcal{L}_{\text{rollout}}
    = \frac{1}{L} \sum_{t=0}^{L-1}
      \bigl\| \hat{\mathbf{s}}_{t+1} - \mathbf{s}_{t+1}^{\text{true}} \bigr\|^2
  \label{eq:rollout-loss}
\end{equation}

\smallskip
\noindent where
\begin{tabular}{@{}>{$}l<{$} @{\quad} l@{}}
  \lambda_r & rollout loss weight ($= 0.1$) \\
  L & rollout length ($= 8$ steps) \\
  \hat{\mathbf{s}}_{t+1} & model's own predicted next state (autoregressive) \\
\end{tabular}

At each step, the model's own prediction $\hat{\mathbf{s}}_{t+1}$ is used as input for the next, and gradients are back-propagated through the full unrolled trajectory.
This loss directly penalizes the drift between surrogate rollouts and ground-truth PyElastica trajectories --- the primary failure mode when the surrogate is used as a substitute environment for RL training.

\subsubsection{Density-Weighted Sampling}
\label{sec:methods:training:density}

Sobol quasi-random action sampling produces a non-uniform distribution over the state space.
To compensate, each training sample is weighted by the inverse of its estimated local density:

\begin{equation}
  w^{(i)} = \frac{1}{\hat{p}(\mathbf{c}^{(i)})}
  \label{eq:density-weight}
\end{equation}

\begin{equation}
  \mathbf{c}^{(i)}
    = \bigl[\bar{x}_{\text{CoM}},\;
            \bar{y}_{\text{CoM}},\;
            \|\dot{\mathbf{x}}\|,\;
            |\bar{\omega}_z|\bigr]
  \label{eq:density-features}
\end{equation}

\smallskip
\noindent where
\begin{tabular}{@{}>{$}l<{$} @{\quad} l@{}}
  w^{(i)} & importance weight for sample $i$ \\
  \hat{p}(\mathbf{c}^{(i)}) & estimated local density of summary features \\
  \bar{x}_{\text{CoM}},\, \bar{y}_{\text{CoM}} & center-of-mass position \\
  \|\dot{\mathbf{x}}\| & center-of-mass speed \\
  |\bar{\omega}_z| & mean angular velocity magnitude \\
\end{tabular}

The density $\hat{p}$ is estimated from a joint histogram with 20 bins per dimension, and weights are clipped at $10\times$ the mean to prevent rare samples from dominating the gradient.

\subsubsection{Hyperparameters}
\label{sec:methods:training:hyperparams}

The best configuration, identified through an architecture sweep (see \cref{sec:discussion:arch}), used: learning rate $1 \times 10^{-3}$ with the Adam optimizer (default momentum parameters), batch size 4096, and early stopping on validation MSE.
All experiments were logged to Weights \& Biases for tracking and comparison.

% ---------------------------------------------------------------------------
\section{Experiments and Results}
\label{sec:experiments}
% ---------------------------------------------------------------------------

\subsection{Surrogate Accuracy}
\label{sec:experiments:accuracy}

\placeholder{Surrogate validation against held-out PyElastica trajectories: single-step MSE, multi-step rollout RMSE, per-state-component accuracy breakdown, and the effect of per-element phase encoding on angular velocity prediction.}

\subsection{RL Training with Surrogate}
\label{sec:experiments:rl}

\placeholder{PPO training in the surrogate environment: learning curves, wall-clock time comparison with direct Elastica training, and policy performance evaluated on both surrogate and real Elastica environments.}

\subsection{Comparison with Direct Elastica RL Baseline}
\label{sec:experiments:baseline}

\placeholder{Head-to-head comparison between surrogate-trained and Elastica-trained policies: sample efficiency, wall-clock time, final policy quality, and the transfer gap attributable to surrogate modeling error.}

% ---------------------------------------------------------------------------
\section{Discussion}
\label{sec:discussion}
% ---------------------------------------------------------------------------

\subsection{Physics Calibration Challenges}
\label{sec:discussion:physics}

Establishing a reliably locomoting Cosserat rod simulation required resolving several categories of physics calibration errors.
Each issue independently prevented stable serpentine locomotion and would have rendered any surrogate trained on the resulting data physically meaningless.
These findings illustrate a general principle: a surrogate model of an incorrectly calibrated simulator will faithfully replicate the simulator's errors.

\textbf{Rod radius and bending stiffness.}
The initial rod radius was set to $r = 0.001$\,m, matching a plausible physical snake scale.
Bending stiffness, however, scales as $EI \propto r^4$: with $I = \pi r^4 / 4$, a Young's modulus of $E = 2 \times 10^6$\,Pa gave $EI \approx 1.6 \times 10^{-6}$\,N$\cdot$m$^2$ --- insufficient to drive meaningful rod deformation.
The rod was effectively limp and produced near-zero locomotion velocity.
Increasing $E$ alone caused the opposite failure: at $E \geq 10^6$\,Pa, the rod snapped violently toward the target curvature, with actual curvatures reaching $\pm 100$\,rad/m (compared to targets of $\pm 2.5$\,rad/m) and node velocities exceeding 178\,m/s, causing simulation divergence.

A grid search over $(E, r)$ pairs identified $r = 0.02$\,m with $E = 10^5$\,Pa as producing stable serpentine motion.
Increasing the radius by a factor of 20 increases $I$ by a factor of $160{,}000$, so even with a simultaneous 20-fold reduction in Young's modulus, the net bending stiffness $EI$ increased sufficiently for physically plausible deformation.
This calibration illustrates a modeling challenge specific to Cosserat rod simulators: the parameter values required to produce correct physical behavior need not correspond to the material parameters of the modeled object, particularly when the simulator's length scales and integration timesteps differ from the physical system by orders of magnitude.

\textbf{Serpenoid wave direction sign error.}
The CPG actuation module contained a sign error in the temporal term of the curvature equation.
The initial implementation computed $\kappa(s, t) = A \sin(2\pi k \cdot s - 2\pi f \cdot t + \phi)$, producing a wave with phase velocity $+f/k$ in the $+s$ direction (tail to head).
Anisotropic RFT friction generates net forward thrust only when the wave propagates posteriorly (head to tail).
The correct expression is $\kappa(s, t) = A \sin(2\pi k \cdot s + 2\pi f \cdot t + \phi)$, with phase velocity $-f/k$ in the $+s$ direction.

This bug was not detected during initial simulator construction because the training environment computed curvatures inline rather than calling the CPG module's export method.
It was identified only after observing consistent negative forward-velocity rewards across multiple training runs, which prompted a comparison of the wave equation against the serpentine gait literature.
The experience motivates explicit unit tests that verify wave propagation direction for a known CPG configuration, rather than relying on training metrics to surface physics errors.

\textbf{RFT friction coefficient tuning.}
Anisotropic drag coefficients derived from published values for soil-crawling snakes ($c_n/c_t \approx 10$) produced excessive lateral drift on the flat-surface simulation substrate.
A grid search over the ratio $c_n/c_t$ found that $c_t = 0.01$ and $c_n = 0.05$ (ratio 5:1) yielded stable forward propulsion with a 94:1 forward-to-lateral velocity ratio across the full CPG parameter range.
The discrepancy from published values reflects the different contact mechanics of dry-surface locomotion (no soil deformation, no granular interlocking), confirming that substrate-specific friction calibration is necessary when transferring the contact model across physical regimes.

\subsection{Surrogate Architecture Experiments}
\label{sec:discussion:arch}

\placeholder{Architecture sweep results: MLP width and depth comparison, residual connections, history window experiments, rollout loss ablation, and the FT-Transformer baseline. Results from Phase 3 Plans 03--05.}

\subsection{Data Collection Pipeline Challenges}
\label{sec:discussion:pipeline}

Four engineering challenges were encountered while building the parallel data collection system, each arising from the interaction between PyElastica's Numba-compiled physics kernels and Python multiprocessing.

\textbf{Numba thread pool deadlock.}
PyElastica's solver is implemented as Numba JIT-compiled functions.
When worker processes are spawned via the \texttt{forkserver} start method, they inherit Python module state but not the Numba thread pool.
On first physics evaluation, Numba re-initializes its thread pool (defaulting to 48 threads on the 48-core machine).
Under heavy load --- 16 workers each spawning 48 threads yields 768 threads competing for 48 cores --- a silent deadlock occurred at episode boundaries.
The fix was to set \texttt{NUMBA\_NUM\_THREADS=1}, eliminating Numba's internal parallelism.
This is the correct trade-off: each PyElastica instance operates on a single 20-element rod, so intra-function parallelism provides no benefit, while process-level parallelism across 16 workers provides a $16\times$ throughput improvement.

\textbf{Stall detection false positives.}
A health monitor tracks per-worker data output rates and respawns stalled workers.
During initial deployment, all 16 workers were flagged as stalled and respawned within 60 seconds of startup.
The cause was that PyElastica initialization --- forking, importing, constructing the rod, and warming the Numba JIT cache --- takes more than 60 seconds under load, exactly matching the stall threshold.
The fix was to condition the stall counter on having produced at least one transition: zero-progress during initialization is not counted.

\textbf{Parallel scaling bottleneck.}
Beyond 16 workers, throughput plateaued rather than continuing to grow linearly.
The bottleneck was memory bandwidth: PyElastica's solver performs many small array operations that are memory-bound, and at high worker counts the aggregate working sets exceed the 30\,MB shared L3 cache per socket.
The optimal configuration was 16 workers (${\sim}8$ per socket), achieving near-linear scaling.
The surrogate ultimately sidesteps this bottleneck entirely: once the dataset is collected, RL training runs on GPU at throughputs orders of magnitude higher than any CPU-bound simulation.

\textbf{Angular velocity coverage gap.}
Analysis of the V1 dataset revealed poor surrogate predictions for element angular velocity ($\omega_z$), with mean R$^2 = 0.23$ and individual elements below zero.
The state perturbation standard deviation of $\sigma_\omega = 0.05$\,rad/s was far below the operational CPG range of 1--10\,rad/s, leaving large-$|\omega_z|$ regions severely underrepresented.
For the V2 dataset, $\sigma_\omega$ was increased to 1.5\,rad/s, covering the full operational range.
This case illustrates a general issue: if the collection distribution does not cover the state-space regions that the downstream RL policy will visit, the surrogate will be inaccurate precisely where accuracy is most needed.

% ---------------------------------------------------------------------------
\section{Conclusion}
\label{sec:conclusion}
% ---------------------------------------------------------------------------

\placeholder{Write conclusion after all experimental results are available.}

% ---------------------------------------------------------------------------
% Bibliography
% ---------------------------------------------------------------------------
\bibliographystyle{plainnat}
\bibliography{references}

\end{document}
